---
output:
  github_document:
    toc: true
    toc_depth: 4
    number_sections: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# riskintrodata

The 'riskintrodata' package is designed to provide a set of functions and datasets
that support the management of data used to estimate the risk of introducing an animal
disease into a specific geographical region.

It includes tools for reading and validating both geographic and tabular datasets
commonly used in the context of animal disease risk estimation.


## Motivation

The primary motivation for creating 'riskintrodata' is to isolate and centralize
the datasets and data import functions required by the 'riskintro' application 
into a dedicated package. This separation simplifies testing, improves clarity,
and makes it easier to document the datasets used in the application in a structured way.

Additionally, the 'riskintrodata' package is designed to simplify package management.
It helps reduce the complexity of handling the numerous packages required by the
'riskintro' application. By centralizing essential datasets and their associated
import functions, 'riskintrodata' minimizes package dependencies.


## Installation

You can install the development version of riskintrodata like so:

```{r eval = FALSE}
pak::pak("ardata-fr/riskintrodata")
```


## Read data

```{r}
library(riskintrodata)
```


The package provides functions to read and validate geographic and tabular datasets.

To read geographic data from a file, you can use the `read_geo_file()` 
function:


```{r}
tun_files <-
  system.file(
    package = "riskintrodata",
    "samples",
    "tunisia",
    "epi_units",
    "tunisia_adm2_clean.gpkg"
  )
read_geo_file(tun_files)

nga_files <- system.file(
  package = "riskintrodata",
  "samples",
  "nigeria",
  "epi_units",
  "NGA-ADM1.geojson"
)
read_geo_file(nga_files)
```

To read raster data, you can use the `read_raster_file()` function:

```{r eval = FALSE}
road_access_raster_file <- download_road_access_raster()
ra_raster <- read_raster_file(x = road_access_raster_file)
ra_raster
```

A third function, `read_emission_risk_factor_file()`, is available to read 
tabular data related to emission risk factors. This function reads a text file
containing emission risk factors for animal diseases. The file should be in a tabular
format with specific columns (see details in the function documentation).

There is no other function to read tabular data, but you can use the
package 'readr' or 'readxl' to read the data if the format is CSV or Excel.

## Validate data

All input datasets used in risk analysis must be validated before they can be processed. The `validate_dataset()` function ensures that datasets meet the expected format and contain the required information for risk calculations.

The validation system checks:

- **Required columns**: Presence of essential fields needed for analysis
- **Data types**: Correct formats for different column types (character, numeric, spatial, etc.)
- **Data values**: Content validation using predefined rules (e.g., valid coordinates, acceptable category values)  
- **Consistency**: Cross-field validation and logical constraints

### Supported dataset types

The function can validate four types of datasets used in risk analysis:

- **Epidemiological units**: `table_name = "epi_units"` - Administrative areas or regions for risk assessment
- **Entry points**: `table_name = "entry_points"` - Border crossings, airports, seaports where animals/products enter  
- **Animal mobility**: `table_name = "animal_mobility"` - Animal movement flows between locations
- **Emission risk factors**: `table_name = "emission_risk_factors"` - Disease control and surveillance measures by country

### Column mapping with the `...` argument  

When your dataset has different column names than expected, use the `...` argument to map your columns to the required field names. This allows you to rename columns during validation without modifying your original dataset.

#### Example: Validating epidemiological units

```{r}
# Load a sample dataset with non-standard column names
nga_files <- system.file(
  package = "riskintrodata", "samples", "nigeria", "epi_units", "NGA-ADM1.geojson"
)
nga_raw <- read_geo_file(nga_files)
colnames(nga_raw)

# Validate by mapping columns: your_column = "required_field"
validated_epi_units <- validate_dataset(
  x = nga_raw,
  table_name = "epi_units", 
  eu_name = "shapeName",     # Map "shapeName" to required "eu_name"
  eu_id = "shapeISO"         # Map "shapeISO" to optional "eu_id"
)

# Extract the clean, validated dataset
clean_epi_units <- extract_dataset(validated_epi_units)
colnames(clean_epi_units)
```

#### Example: Validating entry points

```{r}
# Load sample entry points data  
entry_files <- system.file(
  package = "riskintrodata", "samples", "tunisia", "entry_points", 
  "BORDER_CROSSING_POINTS.csv"
)
entry_raw <- read.csv(entry_files)
colnames(entry_raw)

# Map your columns to required fields
validated_entry_points <- validate_dataset(
  x = entry_raw,
  table_name = "entry_points",
  point_name = "NAME",       # Required: point name
  lng = "LONGITUDE_X",       # Required: longitude  
  lat = "LATITUDE_Y",        # Required: latitude
  mode = "MODE",             # Optional: contraband status
  type = "TYPE"              # Optional: transport type
)

clean_entry_points <- extract_dataset(validated_entry_points)
```

### Validation workflow

1. **Import your data** using `read_geo_file()`, `read_emission_risk_factor_file()`, or standard R functions
2. **Check column names** and identify which fields need mapping
3. **Validate with column mapping** using `validate_dataset()` with appropriate `...` arguments  
4. **Extract clean dataset** using `extract_dataset()` if validation passes
5. **Use in analysis** - validated datasets can be passed directly to risk calculation functions

### Error handling

If validation fails, `validate_dataset()` returns detailed error messages indicating:
- Missing required columns  
- Invalid data types or values
- Specific rows/values that don't meet validation criteria

Use these messages to fix your data and re-validate. Only datasets that pass all validation checks can be used in the risk analysis pipeline.

## Data structures utilities

### Emission risk factors management

`get_wahis_erf` is an helper function for getting the WAHIS emission risk factors dataset. As most analysis done require filtering for one type of each of diease, species and animal_category, this function is a helper for that.

`erf_row` is a function to create a single row of emission risk factors data. It takes parameters corresponding to the columns of the emission risk factors dataset and returns a tibble with the provided values.

`read_emission_risk_factor_file()` reads a text file containing emission risk factors for animal diseases. The file should be in a tabular format with specific columns (see details in the function documentation).

### Building emission risk datasets

Emission risk is calclated from emission risk factors. These factors were originally intended to have come from WAHIS and therefore that dataset is provided in this package. 

A common workflow is to start with WAHIS data for your study parameters and then add custom rows for countries not covered or with updated information.

Note that to calcaulate emission risk score from emission risk factors, you'll need to use the `riskintroanalysis::calc_emission_risk()` function. This uses the default emission risk weights found as shown below.

#### Example: Combining WAHIS data with custom entries

```{r}
library(dplyr)

# Start with WAHIS data for your study parameters
wahis_data <- get_wahis_erf(
  disease = "Anthrax", 
  species = "Cattle", 
  animal_category = "Domestic"
)

# Check what we got from WAHIS
nrow(wahis_data)
head(wahis_data[, c("iso3", "country", "disease", "species", "animal_category")])

# Add custom entries for countries not in WAHIS or with updated information
custom_entry1 <- erf_row(
  iso3 = "XYZ", 
  country = "Example Country",
  disease = "Anthrax",
  animal_category = "Domestic", 
  species = "Cattle",
  disease_notification = 0,        # Good surveillance in place
  targeted_surveillance = 0,       # Active targeted surveillance  
  general_surveillance = 1,        # Limited general surveillance
  screening = 0,                   # Good screening measures
  precautions_at_the_borders = 0,  # Border controls active
  slaughter = 0,                   # Proper slaughter protocols
  selective_killing_and_disposal = 0, # Culling procedures ready
  zoning = 0,                      # Zoning strategies implemented
  official_vaccination = 1,        # No vaccination program
  last_outbreak_end_date = as.Date("2020-01-15"),
  commerce_illegal = 1,            # Some illegal trade suspected
  commerce_legal = 0,              # Legal trade well regulated
  data_source = "Custom entry - Local expert knowledge"
)

# Combine WAHIS data with custom entries
complete_erf <- bind_rows(wahis_data, custom_entry1)

# Verify the combined dataset
nrow(complete_erf)
tail(complete_erf[, c("iso3", "country", "data_source")])

emission_risk_weights
```


## Reference datasets

The package includes several reference datasets that are used in the context of animal disease risk estimation:

- iso3 country codes, available with the function `country_reference()`. 
Also a utility function `iso3_to_name()` is provided to convert ISO3 codes to country names.
- list `emission_risk_weights` contains the emission risk weights by default 
used to calculate emission risk scores and emission risk from emission risk factors.
- `neighbours_table` A correspondence table of all countries and their neighbours,
- `world_sf`, an SF dataset containing global administrative boundaries for most 
countries,
- `wahis_emission_risk_factors`: Emission Risk Factors dataset from WAHIS,
- get default emission risk weights with `get_erf_weights()` function.

