---
output:
  github_document:
    toc: true
    toc_depth: 4
    number_sections: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# riskintrodata

The 'riskintrodata' package provides functions and datasets for managing data used to estimate the risk of introducing an animal disease into a specific geographical region.

## Installation

You can install the development version of riskintrodata like so:

```{r, eval = FALSE}
# Install pak if you don't already have it
install.packages("pak")
pak::pak("git::https://gitlab.cirad.fr/astre/riskintro-app/riskintrodata.git")

# Accept updates to other dependencies
```


## Analysis of introduction risk workflow

This package is the first step to analysing the risk of introduction using the 
riskintroanalysis package. Before analysis can begin, we need to ensure that
the input datasets are validated which makes sure analysis is correct and there 
are no errors.

The overall workflow is: 

1. Import data
2. Validate data
3. Analyse data
4. Visualise data 

The first two steps are done here, by this package. The latter two are done
by riskintroanalysis. There is also the riskintroapp package, a shiny application
that brings all these steps into a graphical user interface.

The first step is reading the data...

## Read data

```{r}
library(riskintrodata)
```

The package provides functions to read and validate geographic and tabular datasets.

1. For geospatial vector datasets (such as adminstrative boundaries) use `read_geo_file()`.

This is most often used for epidemiological units and entry points datasets. 

```{r, eval = FALSE}
tun_files <- system.file(
    package = "riskintrodata",
    "samples", "tunisia", "epi_units", "tunisia_adm2_clean.gpkg"
  )
read_geo_file(tun_files)
```

2. For raster datasets use `read_raster_file()`.

```{r eval = FALSE}
road_access_raster_file <- download_road_access_raster()
ra_raster <- read_raster_file(x = road_access_raster_file)
ra_raster
```

A third function, `read_emission_risk_factor_file()`, is available to read 
tabular data related to emission risk factors. This function reads a csv file
containing emission risk factors for animal diseases. The file should be in a tabular
format with specific columns (see details in the function documentation).

There is no other function to read tabular data, but you can use the
package 'readr' or 'readxl' to read common data formats.

## Validate data

All input datasets used in risk analysis must be validated before they can be processed. The `validate_dataset()` function ensures that datasets meet the expected format and contain the required information for risk calculations.

The validation system checks:

- Required columns: Presence of essential fields needed for analysis
- Data types: Correct formats for different column types (character, numeric, spatial, etc.)
- Data values: Content validation using predefined rules (e.g., valid coordinates, acceptable category values)  

If your dataset isn't valid, don't worry it will give you the details why.

### Supported dataset types

The function `validate_dataset()` validates the 4 input datasets.

- Epidemiological units: `table_name = "epi_units"` - Administrative areas or regions for risk assessment. 
Used in all riskintroanalysis analyis methods.

- Entry points: `table_name = "entry_points"` - Border crossings, airports, seaports where animals/products enter. Used in riskintroanalysis entry points analysis.

- Animal mobility: `table_name = "animal_mobility"` - Animal movement flows between locations. 
Used in riskintroanalysis animal mobility analysis.

- Emission risk factors: `table_name = "emission_risk_factors"` - Disease control and surveillance measures by country. Used in entry points, border risk and animal mobility introduction risk analysis.

### Column mapping with the `...` argument  

When your dataset has different column names than those that are required by `validate_dataset()`, use the `...` argument to map your columns to the required field names. Alternatively, you can rename
your columns to match data requirements, which case you don't need to provide `...` arguments.

See the following code blocks for examples.

#### Example: Validating epidemiological units

```{r}
# Load a sample dataset with non-standard column names
nga_files <- system.file(
  package = "riskintrodata", "samples", "nigeria", "epi_units", "NGA-ADM1.geojson"
)
nga_raw <- read_geo_file(nga_files)
colnames(nga_raw)

# Validate by mapping columns: your_column = "required_field"
validated_epi_units <- validate_dataset(
  x = nga_raw,
  table_name = "epi_units", 
  eu_name = "shapeName",     # Map "shapeName" to required "eu_name"
  eu_id = "shapeISO"         # Map "shapeISO" to optional "eu_id"
)

# Extract the clean, validated dataset
clean_epi_units <- extract_dataset(validated_epi_units)
clean_epi_units
```

#### Example: Validating entry points

```{r}
# Load sample entry points data  
entry_files <- system.file(
  package = "riskintrodata", "samples", "tunisia", "entry_points", 
  "BORDER_CROSSING_POINTS.csv"
)
entry_raw <- read.csv(entry_files)
colnames(entry_raw)

# Map your columns to required fields
validated_entry_points <- validate_dataset(
  x = entry_raw,
  table_name = "entry_points",
  point_name = "NAME",       # Required: point name
  lng = "LONGITUDE_X",       # Required: longitude  
  lat = "LATITUDE_Y",        # Required: latitude
  mode = "MODE",             # Optional: contraband status
  type = "TYPE"              # Optional: transport type
)

clean_entry_points <- extract_dataset(validated_entry_points)
clean_entry_points
```

### Validation workflow

1. **Import your data** using `read_geo_file()`, `read_emission_risk_factor_file()`, or other R functions and packages.
2. **Check column names** and identify which fields need mapping
3. **Validate with column mapping** using `validate_dataset()` with appropriate `...` arguments  
4. **Extract clean dataset** using `extract_dataset()` if validation passes
5. **Use in analysis** - validated datasets can be passed directly to risk calculation functions

### Error handling

If validation fails, `extract_dataset()` returns detailed error messages indicating:
- Missing required columns  
- Invalid data types or values
- Specific rows/values that don't meet validation criteria

Use these messages to fix your data and re-validate. Only datasets that pass all validation checks can be used in the risk analysis workflow.

## Data structures utilities

### Emission risk factors management

`get_wahis_erf` gives access the WAHIS emission risk factors dataset. Introduction
risk analysis is done for one animal disease, one species, and one animal category at a time.
Use the arguments `disease`, `species` and `animal_category` to choose.

```{r}
# Start with WAHIS data for your study parameters
wahis_data <- get_wahis_erf(
  disease = "Anthrax", 
  species = "Cattle", 
  animal_category = "Domestic"
)
```

`erf_row` is a function to create a single row of emission risk factors data. It takes parameters corresponding to the columns of the emission risk factors dataset and returns a tibble with the provided values.

This is useful when you want to update or add rows to the WAHIS data. Use dplyr functions
such `rows_upsert` to update and insert rows.

```{r}
library(dplyr)
custom_entry1 <- erf_row(
  iso3 = "XYZ", 
  country = "Example Country",
  disease = "Anthrax",
  animal_category = "Domestic", 
  species = "Cattle",
  disease_notification = 0,        # Good surveillance in place
  targeted_surveillance = 0,       # Active targeted surveillance  
  general_surveillance = 1,        # Limited general surveillance
  screening = 0,                   # Good screening measures
  precautions_at_the_borders = 0,  # Border controls active
  slaughter = 0,                   # Proper slaughter protocols
  selective_killing_and_disposal = 0, # Culling procedures ready
  zoning = 0,                      # Zoning strategies implemented
  official_vaccination = 1,        # No vaccination program
  last_outbreak_end_date = as.Date("2020-01-15"),
  commerce_illegal = 1,            # Some illegal trade suspected
  commerce_legal = 0,              # Legal trade well regulated
  data_source = "Custom entry - Local expert knowledge"
)
complete_erf <- dplyr::rows_upsert(wahis_data, custom_entry1, by = "iso3") |> 
  arrange(desc(iso3))
complete_erf
```




Incase you have a premade dataset you can use the `read_emission_risk_factor_file()` reads a text file containing emission risk factors for animal diseases. The file should be in a tabular format with specific columns (see details in the function documentation). You can also validate emission risk factors with `validate_dataset(table_name = "emission_risk_factors",...)`.

## Reference datasets

The package includes several reference datasets that are used in the context of animal disease risk estimation:

- iso3 country codes, available with the function `country_reference()`.
Also a utility function `iso3_to_name()` is provided to convert ISO3 codes to country names.
- list `emission_risk_weights` contains the emission risk weights by default
used to calculate emission risk scores and emission risk from emission risk factors.
- `neighbours_table` A correspondence table of all countries and their neighbours,
- `world_sf`, an SF dataset containing global administrative boundaries for most
countries,
- `wahis_emission_risk_factors`: Emission Risk Factors dataset from WAHIS,
- get default emission risk weights with `get_erf_weights()` function.

## Package motivation

The primary motivation for creating 'riskintrodata' is to isolate and centralize
the datasets and data import functions required by the 'riskintro' application
into a dedicated package. This separation simplifies testing, improves clarity,
and makes it easier to document the datasets used in the application in a structured way.

Additionally, the 'riskintrodata' package is designed to simplify package management.
It helps reduce the complexity of handling the numerous packages required by the
'riskintro' application. By centralizing essential datasets and their associated
import functions, 'riskintrodata' minimizes package dependencies.

